---
title: "Macrophage GxE study QC report"
output:
  html_document:
    keep_md: yes
---
  
```{r echo=FALSE}
# This chunk contains some setup which we do not want to display in the final
# document, hence `echo=FALSE` is used.
# When you’re reading this document, it may be easier to ignore this chunk for
# now, and continue with the document below, starting at “Introduction”
 
# Make library loading silent
library = function (...) suppressMessages(base::library(...))
library("knitr")
opts_knit$set(root.dir = '../../../')
library("ggplot2")
library("gplots")
library("devtools")
library("dplyr")
load_all("../../../../seqUtils/")
load_all("../seqUtils/")
```

Load eQTL dataset from disk and extract peer factors for different conditions
```{r}
data_list = readRDS("results/SL1344/eqtl_data_list.rds")
naive_peer = data_list$covariates_list[[1]][4:13,]
IFNg_peer = data_list$covariates_list[[2]][4:13,]
SL1344_peer = data_list$covariates_list[[3]][4:13,]
IFNg_SL1344_peer = data_list$covariates_list[[4]][4:13,]

```

```
Explore the correlation between PEER factors
```{r, fig.width = 8, fig.height = 8}
heatmap.2(abs(cor(t(naive_peer))), margins = c(10,10))
heatmap.2(abs(cor(t(SL1344_peer))), margins = c(10,10))
heatmap.2(abs(cor(t(IFNg_SL1344_peer))), margins = c(10,10))
heatmap.2(abs(cor(t(IFNg_peer))), margins = c(10,10))

```

Identify which factors are most correlated with PEER factors
```{r,fig.width = 8, fig.height = 6}
covariates = dplyr::select(expanded_meta, mean_purity_filtered, ng_ul_mean, RNAseq_auto, diff_days, passage_diff)

peer_explained = explainPEER(peer_factors[,1:5], covariates)
peer_explained$peer_factor = factor(peer_explained$peer_factor, 
                                    levels = paste("PEER_factor_", c(1:ncol(peer_factors)), sep = ""))
ggplot(peer_explained, aes(x = peer_factor, y = covariate, fill = sqrt(r_squared))) + 
  geom_tile() + 
  scale_fill_gradient2(space = "Lab", low = scales::muted("blue"), 
                       high = scales::muted("red"), midpoint = 0.4, name = "Correlation")
```

Make the same plot by excluding some of the purity outliers
```{r, fig.width = 8, fig.height = 6}
filter = expanded_meta$mean_purity_filtered > 0.95
peer_explained = explainPEER(peer_factors[filter,1:5], covariates[filter,])
peer_explained$peer_factor = factor(peer_explained$peer_factor, 
                                    levels = paste("PEER_factor_", c(1:ncol(peer_factors)), sep = ""))
ggplot(peer_explained, aes(x = peer_factor, y = covariate, fill = sqrt(r_squared))) + 
  geom_tile() + 
  scale_fill_gradient2(space = "Lab", low = scales::muted("blue"), 
                       high = scales::muted("red"), midpoint = 0.4, name = "Correlation")
```

Visualize the RNA-Seq batch effect,
```{r,fig.width = 8, fig.height = 6}
ggplot(expanded_meta, aes(x = PEER_factor_1, y = PEER_factor_2, label = sample_id, color = RNAseq_auto)) + 
  geom_point() + geom_text()
```

Visualize the purity effect with one outlier,
```{r,fig.width = 8, fig.height = 6}
ggplot(expanded_meta, aes(x = mean_purity_filtered, y = PEER_factor_1, label = sample_id)) + 
  geom_point() + geom_text()
```

Remove bubh outlier,
```{r,fig.width = 8, fig.height = 6}
ggplot(dplyr::filter(expanded_meta, mean_purity_filtered > 0.9), aes(x = mean_purity_filtered, y = PEER_factor_1, label = sample_id)) + 
  geom_point() + geom_text()
```

Correlation between factor 1 and RNA concentration
```{r,fig.width = 8, fig.height = 6}
ggplot(expanded_meta, aes(x = ng_ul_mean, y = PEER_factor_1, label = sample_id)) + geom_point() + geom_text()
cor.test(as.numeric(factor(expanded_meta$ng_ul_mean)), expanded_meta$PEER_factor_1, method = "pearson")
```

#Factors detected in Salmonella condition
```{r}
precisions = read.table("results/SL1344/PEER/25_factors_cond_C/precision.txt", sep = ",")
precisions_vec = precisions$V1[2:nrow(precisions)]
plot(1/precisions_vec)
```

Import PEER results
```{r}
peer_factors = read.table("results/SL1344/PEER/25_factors_cond_C/factors.txt", sep =",")
peer_factors = peer_factors[,2:ncol(peer_factors)]
colnames(peer_factors) = paste("PEER_factor_", c(1:ncol(peer_factors)), sep = "")
```

Add PEER results to the sample metadata
```{r}
peer_factors_2 = dplyr::mutate(peer_factors, sample_id = line_meta$sample_id)
expanded_meta = dplyr::left_join(line_meta, peer_factors_2, by = "sample_id")
expanded_meta = dplyr::mutate(expanded_meta, RNAseq_auto = ifelse(chemistry == "V4_auto", "yes", "no"))
```

ggplot(expanded_meta, aes(x = PEER_factor_1, y = PEER_factor_2, label = sample_id, color = RNAseq_auto)) + 
  geom_point() + geom_text()
  
Remove bubh outlier,
```{r,fig.width = 8, fig.height = 6}
ggplot(dplyr::filter(expanded_meta, mean_purity_filtered > 0.9), aes(x = mean_purity_filtered, y = PEER_factor_1, label = sample_id)) + 
  geom_point() + geom_text()
```

```{r,fig.width = 8, fig.height = 6}
ggplot(expanded_meta, aes(x = RNAseq_auto, y = PEER_factor_3, label = sample_id, color = RNAseq_auto)) + 
  geom_point() + geom_text()
```

