---
title: "Variance Component analysis of Salmonella RNA-Seq data"
output: html_document
---

```{r echo=FALSE}
# This chunk contains some setup which we do not want to display in the final
# document, hence `echo=FALSE` is used.
# When you’re reading this document, it may be easier to ignore this chunk for
# now, and continue with the document below, starting at “Introduction”

# Make library loading silent
library = function (...) suppressMessages(base::library(...))
options(warn=-1)
project_home = "../../../"
library("dplyr")
library("devtools")
library("ggplot2")
load_all(paste(project_home, "macrophage-gxe-study/seqUtils/", sep = ""))
```
The varianceComponents.R script generated two tables that contain the variance explaind by each factor for each gene. Let's look at the compact model first.

First, lets import the variance compoment estimates for each gene:
```{r}
model_compact = readRDS(paste(project_home, "results/varComp/model_compact_results.rds", sep = "")) %>% tbl_df() %>%
  dplyr::filter(converged == TRUE) %>% #Remove genes where lmr4 did not converge
  dplyr::select(-type, -converged) %>% #Remove unnecessary columns
  dplyr::rename(residual = Residual,SL1344_IFNg = `SL1344:IFNg`, batch = salmonella) #Rename some factos
```

Now, we first bin the genes by the amount of residual variance. We notice that only a small number of genes have high residual variance. 
```{r}
binned_table = binGenesByResidual(model_compact, n_bins = 20)
bin_size_table = dplyr::group_by(binned_table, residual_bin) %>% dplyr::summarise(bin_size = length(residual_bin))
bin_sizes_plot = ggplot(bin_size_table, aes(x = residual_bin, y = bin_size)) + geom_bar(stat = "identity") + 
  ylab("Number of genes") + 
  xlab("Residual variance bin")
bin_sizes_plot
```

We can now look what are the factors that explain variance in each bin. 

```{r}
var_explained = meanVarianceWithinBins(binned_table)
plotBinnedVariance(var_explained)
```

However, this plot becomes hard to read, becuase there are so many different factors. Let's focus on the total variance explaind by each component. Here is the mean variance explained by each factor:

```{r}
dplyr::filter(var_explained, residual_bin == "Total") %>% 
  arrange(-var_explained)
```

We can also look at the distribution of variance explained across genes:

```{r, fig.width=10}
dat = tidyr::gather(model_compact, factor, var_explained, donor:SL1344_IFNg)
ggplot(dat, aes(x = factor, y = var_explained)) + geom_violin(scale = "width" ) + 
  geom_boxplot(width = .2, outlier.shape = NA)
```

Or, for each factor, we count the number of genes where this particular factor explains the most variance:

```{r}
maximum_factor = maximumFactorPerGene(model_compact)
table(maximum_factor$component_max)
```
### Pooled model
Since we noticed that library pool and rna concentration explain only a small amout of the variation we can pool these together with batch and name the derived factor as technical variance.

```{r}
model_pooled = dplyr::transmute(model_compact, gene_id, donor, gender, IFNg, SL1344, SL1344_IFNg, residual, 
                                technical = batch + library_pool + rna_concentration, purity = purity_bins)
binned_pooled = binGenesByResidual(model_pooled, n_bins = 20)
var_pooled = meanVarianceWithinBins(binned_pooled)
var_pooled_plot = plotBinnedVariance(var_pooled)
var_pooled_plot
```

### Expaned model
The extended model contains three additional factors: passage of the iPS cells prior to differenitation, duration of differntiation from iPS cells to macrophages and duration of differentiation from progenitors to macrophages:

```{r}
model_extended = readRDS(paste(project_home, "results/varComp/model_extended_results.rds", sep = "")) %>% tbl_df() %>%
  dplyr::filter(converged == TRUE) %>% #Remove genes where lmr4 did not converge
  dplyr::select(-type, -converged) %>% #Remove unnecessary columns
  dplyr::rename(residual = Residual,SL1344_IFNg = `SL1344:IFNg`, batch = salmonella) #Rename some factos
```

It seems that neither the duration of macrophage differentiation (diff_bins) nor the the passage of the iPS line prior to differentiation seem to explain large proportion of the variance. Days of macrophage differntiation (6,7 or 8) seems to explain some variance, but this can be confounded by batch, because cells in the sme batch were always treated on the same day. Also the amount of variance explained by purtiy_bins seems quite high, but we need more data to be condifent about that, because 4/7 purity low purity samples are from the first stimulation batch.

```{r}
extended_binned = binGenesByResidual(model_extended, n_bins = 20)
extended_var_explained = meanVarianceWithinBins(extended_binned)
dplyr::filter(extended_var_explained, residual_bin == "Total") %>% 
  arrange(-var_explained)
```
For completeness, here is the violin plot:
```{r, fig.width=10}
dat = tidyr::gather(model_extended, factor, var_explained, donor:SL1344_IFNg)
ggplot(dat, aes(x = factor, y = var_explained)) + geom_violin(scale = "width" ) + 
  geom_boxplot(width = .2, outlier.shape = NA)
```
